---
# output: github_document
title: Error detection in science
output:
  rmarkdown::md_document:

vignette: >
  %\VignetteIndexEntry{Error detection in science}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{devtools}

# bibliography: references.bib
link-citations: TRUE
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  fig.width = 4.5,
  fig.asp = 0.8
)

# Dev only: load scrutiny from scrutiny
devtools::load_all(".")

# Determine ggplot2 theme for the session:
ggplot2::theme_minimal() %>% 
  ggplot2::theme_set()
```

<!-- badges: start -->

<!-- badges: end -->

The goal of scrutiny is to test published summary statistics for consistency and reconstruct some of the processes which created them. The package makes these methods easy to use in a tidyverse-friendly way. It hopes to help the new field of error detection go mainstream.

Besides consistency tests, scrutiny features infrastructure for implementing error detection techniques, as well as specialized data wrangling functions. See the *Articles* tab for vignettes.

<!-- ## Installation -->

<!-- You can install the development version of scrutiny like so: -->

<!-- ``` r -->

<!-- # FILL THIS IN! HOW CAN PEOPLE INSTALL YOUR DEV PACKAGE? -->

<!-- ``` -->

<!-- ## Example -->

Here is how to GRIM-test all values in a data frame. When using `grim_map()`, the `consistency` column tells you if the means (`x`) and sample sizes (`n`) are mutually consistent.

```{r example}
library(scrutiny)

# Example data:
pigs1

# GRIM-testing for data frames:
grim_map(pigs1)
```

Test percentages instead of means:

```{r}
pigs2

grim_map(pigs2, percent = TRUE)
```

Visualize GRIM results while also selecting a rounding procedure:

```{r fig.width=4, fig.height=4}
pigs1 %>% 
  grim_map(rounding = "up") %>% 
  grim_plot()
```

Similarly, test means and standard deviations of binary data with DEBIT:

```{r fig.width=4.5, fig.height=4}
pigs3

pigs3 %>% 
  debit_map()

pigs3 %>% 
  debit_map() %>% 
  debit_plot()
```

# Guiding ideas

> (...) a critical inspection of the published literature should not be mischaracterized as a hobby for the overly cynical, nor as so-called "methodological terrorism". On the contrary, carefully evaluating presented data is a cornerstone of scientific investigation, and it is only logical to apply this also to the published literature. If we are not willing to critically assess published studies, we also cannot guarantee their veracity.

--- van der Zee et al. (2017, pp. 8-9)

> (...) data thugs (...) demand data and if they do not receive it, they contact editors and universities and threaten to write blogs and tweets about the errors uncovered.

--- Eric A. Stewart (six retractions; quoted in Pickett 2020, p. 178)

# References

Pickett, Justin T. „The Stewart Retractions: A Quantitative and Qualitative Analysis". *Econ Journal Watch* 17, Nr. 1 (March 2020): 152–90. URL: <https://econjwatch.org/articles/the-stewart-retractions-a-quantitative-and-qualitative-analysis>.

van der Zee, Tim, Jordan Anaya, und Nicholas J. L. Brown. „Statistical Heartburn: An Attempt to Digest Four Pizza Publications from the Cornell Food and Brand Lab". *BMC Nutrition* 3, Nr. 1 (Dezember 2017): 54. <https://doi.org/10.1186/s40795-017-0167-x>.
