---
title: "Duplication analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Duplication analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r include=FALSE}
# Dev only: load scrutiny from within scrutiny
devtools::load_all(".")
```

```{r setup}
library(scrutiny)
```

You can use scrutiny to analyze duplicate values in data. Duplications can go a long way in assessing the reliability of published research.

This vignette walks you through scrutiny's tools for finding, counting, and summarizing duplications. It uses the `pigs4` dataset as a simple example:

```{r}
pigs4
```

## Ranked counting with `duplicate_count()`

A good first step is to create a frequency table. To do so, use `duplicate_count()`:

```{r}
pigs4 %>% 
  duplicate_count()
```

It returns a tibble (data frame) that lists all unique `value`s. It is ordered by the `count` of values in the input data frame, so the values that appear most often are at the top. The `locations` are the names of the column or columns in which a given value appears. They are counted by `locations_n`.

For larger datasets, summary statistics can be helpful. Just run `audit()` after `duplicate_count()`:

```{r}
pigs4 %>% 
    duplicate_count() %>% 
    audit()
```

## Counting by column pair with `duplicate_count_colpair()`

Sometimes, a sequence of data may be repeated in multiple columns. `duplicate_count_colpair()` helps find such cases:

```{r}
pigs4 %>% 
  duplicate_count_colpair()
```

`x` and `y` represent all combinations of columns in `pigs4`. The `count` is the number of values that appear in both respective columns. This is different from `duplicate_count()`, where `count` displays total frequencies.

`snout` and `tail` are the column pair with the most overlap: 2 out of 5 values are the same, a rate of 0.4. If there are no missing values, `total_x` and `total_y` are the same. The same applies to `rate_x` and `rate_y`.

Again, you can get summary statistics with `audit()`:

```{r}
pigs4 %>% 
  duplicate_count_colpair() %>% 
  audit()
```

## Counting by observation with `duplicate_tally()`

Unlike the other two functions, `duplicate_tally()` preserves the structure of the original data frame. It adds an `_n` column next to each original column. The newly added columns count how often each value appears in the data frame as a whole:

```{r}
pigs4 %>% 
    duplicate_tally()
```

In `snout`, for example, `4.22` appears twice, so its entries in `snout_n` are `2`. But likewise, `8.13` appears in both `snout` and `tail`, so both observations are marked `2` in the `_n` columns.

When following `duplicate_tally()` up with `audit()`, it shows summary statistics for each `_n` column. The last row summarizes all of these columns together.

```{r}
pigs4 %>% 
    duplicate_tally() %>% 
    audit()
```
