---
title: "Infrastructure beyond rounding"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Infrastructure beyond rounding}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{devtools}
bibliography: references.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r include=FALSE}
# Dev only: load scrutiny from within scrutiny
devtools::load_all(".")
```

```{r setup, message=FALSE}
library(scrutiny)
```

The implementation of error detection techniques in scrutiny rests on a foundation of specialized helper functions. Some of these are exported because they might be helpful in error detection more broadly, or perhaps even in other contexts.

This vignette provides an overview of scrutiny's miscellaneous infrastructure for implementing error detection techniques. It does not discuss infrastructure for rounding, a topic so important to the package that it has its own article: `vignette("rounding")`.

## Count decimal places

Large parts of the package ultimately rest on either of two functions that simply count decimal places. These are digits after a number's decimal point or some other separator. Both functions also take strings.

`decimal_places()` is vectorized:

```{r}
decimal_places("2.80")

decimal_places(c(55.1, 6.493, 8))

vec1 <- iris %>% 
  dplyr::slice(1:10) %>% 
  dplyr::pull(Sepal.Length)

vec1

vec1 %>% 
  decimal_places()
```

Using strings (that are coercible to numeric) is recommended in an error detection context because trailing zeros can be crucial here. Numeric values drop trailing zeros, whereas strings preserve them:

```{r}
decimal_places(7.200)

decimal_places("7.200")
```

`decimal_places_scalar()` is considerably faster than `decimal_places()` but only takes a single number or string into account. This makes it suitable as a helper within other single-case functions.

## Restore trailing zeros

When dealing with numbers that used to have trailing zeros but lost them from being registered as numeric, call `restore_zeros()` to format them correctly. This can be relevant within functions that create vectors where trailing zeros matter, such as the `seq_*()` functions presented in the next section.

Suppose all of the following numbers originally had one decimal place, but some no longer do:

```{r}
vec <- c(4, 6.9, 5, 4.2, 4.8, 7, 4)

vec %>% 
  decimal_places()
```

Now, get them back with `restore_zeros()`:

```{r}
vec %>% 
  restore_zeros()

vec %>% 
  restore_zeros() %>% 
  decimal_places()
```

This uses the default of going by the longest mantissa and padding the other strings with decimal zeros until they have that many decimal places. However, this is just a heuristic: The longest mantissa might itself have lost decimal places. Specify the `width` argument to explicitly state the desired mantissa length:

```{r}
vec %>% 
  restore_zeros(width = 2)

vec %>% 
  restore_zeros(width = 2) %>% 
  decimal_places()
```

## Sequence generation

### The limits of `seq()`

`base::seq()` is an elegant way to generate regular sequences, but it is almost too elegant: The way to determine the step size entirely depends on which arguments are added with non-standard evaluation (i.e., via the dots, `…`).

By default, the step size is assumed to be 1, which can be overridden using the `by` parameter. However, `by` clashes with the only way to state the desired output length, which is the `length.out` parameter. This makes the step size dependent on `from`, `to`, and `by`, and therefore highly variable in the face of different specifications of these parameters. For example:

```{r}
seq(from = 4, to = 7, length.out = 5)

seq(from = 4, to = 20, length.out = 5)
```

It is worth emphasizing that `seq()` encapsulates two mutually exclusive ways of generating sequences: (1) by a manually specified step size without controlling output length; and (2) by a manually specified output length without controlling step size. These two approaches overdetermine each other, and are therefore irreconcilable. Here's what you get if you try it both ways within a single call:

```{r error=TRUE}
seq(from = 4, to = 7, by = 0.1, length.out = 5)
```

Another issue is that `seq()` doesn't automatically take the input numbers' decimal depth into account (i.e., unless `by` was specified in such a way), which would be desirable in an error detection context.

With regards to decimal sequences, therefore, `seq()` does both too much and too little. It has two distinct functional rationales that uneasily sit together in a single function. One might wish to disentangle them: to generate sequences that either run from A to B at the appropriate decimal level, or that have some given length without jeopardizing the step size. It might even be worth spelling out the two rationales in distinct functions.

As it happens, scrutiny features two wrappers around `seq()` that separately capture both of its modi operandi, all while guided by the input numbers' decimal depth: `seq_endpoint()` and `seq_distance()`. Both functions forsake the `by` argument, but that's actually good because they infer the step size automatically from the number of decimal places in their input values.

The package also has two functions that build up on these basic ones by quickly and flexibly constructing data frames with the respective kind of sequence, `seq_endpoint_df()` and `seq_distance_df()`. These latter two accept new columns entered as in `tibble::tibble()` or `dplyr::mutate()`.

Here, I will first present the two basic `seq()` wrappers. Next, I will briefly discuss the higher-level `*_df` functions.

### Two `seq()` wrappers

#### Sequence between two fix points

Consider this call to `seq()`:

```{r}
seq(from = 4.1, to = 7)
```

The intended output of such a call is likely a sequence that progresses in steps of 0.1, and that actually arrives at 7. `base::seq()` won't understand this, but `scrutiny::seq_endpoint()` will:

```{r}
seq_endpoint(from = 4.1, to = 7)
```

Unlike `seq()`, therefore, `seq_endpoint()` always reaches its target. The last value in a vector returned by `seq_endpoint()` will be numerically equal to the value of its `to` argument, no matter how the arguments were specified.

(The output is string by default, but it doesn't need to be:)

```{r}
seq_endpoint(from = 4.1, to = 7, string_output = FALSE)
```

For step size ($S$), the function counts decimal places in `from` and `to`, and then looks for the greater number of decimal places ($D_{max}$). It proceeds by one unit of the longer mantissa's last decimal digit:

$$
S = \frac{1}{10^{D_{max}}}
$$

So if the endpoint has more decimal places than the starting point, it works like this:

```{r}
seq_endpoint(from = 4.5, to = 4.62)
```

Because the step size in `seq_endpoint()` is determined by the greater decimal depth from among its main arguments, it's possible to set off the starting point and/or the endpoint of the sequence at that very level:

```{r}
seq_endpoint(from = 0, to = 0.5)                    # no offset

seq_endpoint(from = 0, to = 0.5, offset_from = 3)   # start at 0.3 instead of 0

seq_endpoint(from = 0, to = 0.5, offset_to = 3)     # end at 0.8 instead of 0.5
```

It also works with negative offset values:

```{r}
seq_endpoint(from = 0, to = 0.5, offset_from = -3)   # start at -0.3 instead of 0

seq_endpoint(from = 0, to = 0.5, offset_to = -3)     # end at 0.2 instead of 0.5
```

#### Sequence of stated length

`seq_distance()` doesn't take a `to` argument --- it only takes a starting point and generates a sequence from there. The length can be chosen at will, and the sequence always proceeds by one unit of `from`'s last decimal digit:

```{r}
seq_distance(from = 8, length_out = 5)

seq_distance(from = 3.2, length_out = 5)
```

The default length is 10:

```{r}
seq_distance(29.97)
```

Reverse the direction with `dir = -1`:

```{r}
seq_distance(from = 2.7, dir = -1)
```

Offsetting the starting point works as in `seq_endpoint()`, although there is no `offset_to` argument:

```{r}
seq_distance(from = 2.7, offset_from = 4)

seq_distance(from = 2.7, offset_from = -4)
```

### Data frames with sequence columns

Both `seq_*` functions from above have a `seq_*_df()` variant. These are meant for interactive analysis, e.g., GRIM-testing the neighbors of an inconsistent value. New columns can be created within the call as in `tibble::tibble()` or `dplyr::mutate()`. All arguments from the basic functions are available, but with a dot before each.

Output of these functions has the correct format for scrutiny's data frame testing functions like `grim_map()`. See `vignette("grim")`, section *Testing numeric sequences*.

Here is the endpoint data frame variant, with an extra column added:

```{r}
seq_endpoint_df(.from = "4.50", .to = 4.57, n = 30)
```

And here is the distance data frame variant:

```{r}
seq_distance_df(.from = "4.50", .length_out = 5, n = 30)
```

## Vary hypothetical group sizes

### The problem

The reporting of summary statistics is often insufficient --- certainly from an error detection point of view. In particular, values such as means and standard deviations are not always accompanied by their respective group sizes, but only by a total sample size.

This presents a problem for consistency tests that rely on reported group sizes, such as GRIM. It requires splitting the reported total into groups and creating multiple plausible scenarios of group sizes that add up to the total. Although no definitive test results can be gained this way, it does help to see whether reported values are consistent with at least some of the plausible group sizes [@bauer_expression_2021].

### Producing functions with `function_map_disperse()`

For GRIM, this problem can be tackled with `grim_map_disperse()`. See the [case study in the GRIM vignette](https://lhdjung.github.io/scrutiny/articles/grim.html#handling-unknown-group-sizes-with-grim_map_disperse) for how its logic works out in practice. Although the underlying code is fairly complex, the function itself was written in an extremely simple way --- by calling `function_map_disperse()`:

```{r}
grim_map_disperse <- function_map_disperse(
  .fun = grim_map, .name_test = "GRIM"
)
```

The same pattern can be applied to any function like `grim_map()`, i.e., a function that tests the values contained in two columns of a data frame for consistency. One of the columns, `n`, will have its values dispersed from half using `disperse_total()`, explained below. It's not currently possible to include more than two columns in this scheme (prohibiting use for DEBIT), but that might change in the future.

To export a function created this way from your own package, make sure to follow [this purrr FAQ](https://purrr.tidyverse.org/reference/faq-adverbs-export.html). (Incredible as it sounds, scrutiny will then take on the role of purrr.) Also, see this [worked example](https://lhdjung.github.io/scrutiny/reference/function_map_disperse.html#ref-examples) for the `.onLoad` pattern.

### Dispersal from (around) half with `disperse_total()`

Briefly, `disperse_total()` checks if an input total is even or odd, cuts it in half, and creates "dispersed" group sizes going out from there, with each pair of group sizes adding up to the input total. This works naturally with even totals. For odd totals, it starts with the two integers closest to half.

The function internally calls either of `disperse()` and `disperse2()`, but I recommend simply using the higher-level `disperse_total()`. Here are two basic examples:

```{r}
# With an even total...
disperse_total(n = 70)

# ...and with an odd total:
disperse_total(n = 83)
```

## Test for subsets, supersets, and equal sets

Starting with `is_subset_of()`, scrutiny features a distinctive family of predicate functions that test whether one vector `x` is a subset of another vector `y`, whether `x` is a superset of `y` (i.e. the reverse of a subset), or whether `x` and `y` are equal sets.

As a teaser: These functions are divided into three subgroups based on the way the second vector, `y`, is constituted. For example, you might test if `x` is a subset of multiple other vectors taken together, or a superset of a vector `y` that consists of multiple values entered along with `x`.

Functions from this family are not currently used as helpers inside other scrutiny functions, but that may well change. Use elsewhere is also conceivable.

# References
