---
title: "Rounding options"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Rounding options}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message=FALSE}
library(scrutiny)
```

Many scrutiny functions have a `rounding` argument that controls how numbers are assumed to have been rounded. Specify `rounding` as any of the strings below.

## Start here: you usually don't need to know

For consistency testing, **the default is almost always the right choice**, and identifying the software behind a paper is usually unnecessary.

`rounding = "up_or_down"` counts a value set as consistent if the reported number matches *either* of the two results. Its range therefore spans every way of breaking a tie at 5 --- up, down, to even, away from zero, toward zero --- so it cannot reject a value set on the grounds of a tie-breaking rule you guessed wrong. `rounding = "ceiling_or_floor"` is wider still and covers every rounding procedure whatsoever, including truncation and custom thresholds.

Committing to a single procedure (`"up"`, `"down"`, `"even"`, `"trunc"`, ...) only ever makes a test **stricter**. That is worth doing when you have positive evidence about how the numbers were produced --- and it is a way to get a false positive when you don't. If in doubt, leave `rounding` alone.

The rest of this vignette is for the case where you do have such evidence.

## In scrutiny

These options return two rounded numbers for each input number. In consistency tests, this means counting a value set as consistent if the reported number matches either of the two rounded reconstructed numbers.

-   `"up_or_down"` (the default): one number rounded up from 5, one down from 5; in this order.
-   `"up_from_or_down_from"`: one number rounded up, one down --- but not from 5. Instead, specify `threshold` as the number from which the input should be rounded.
-   `"ceiling_or_floor"`: one number ceiled, one floored.

By contrast, these options only return one rounded number per input:

-   `"even"` rounds to the next even number, using base R's own `round()`. See the note below.
-   `"up"` or `"down"` rounds in the specified direction, starting at 5.
-   `"up_from"` or `"down_from"` rounds in the specified direction, starting at some number other than 5. Specify this number via the `threshold` argument.
-   `"ceiling"` or `"floor"` always rounds to the next higher or lower decimal place, respectively.
-   `"trunc"` and `"anti_trunc"` always round towards zero or away from it.

Two arguments modify these procedures:

-   `symmetric` decides how ties in **negative** numbers are broken. It is not a technicality --- see the next section.
-   `threshold` moves the point at which `"up_from"` and `"down_from"` switch direction. See `round_up_from()`.

Finally, four options name a complete tie-breaking procedure, so that `symmetric` does not have to be given separately (and is ignored if it is):

-   `"ties_up"` sends a tie to the higher number, `"ties_down"` to the lower one.
-   `"ties_away"` sends a tie away from zero, `"ties_zero"` toward it.

`"ties_up"` is `"up"` with `symmetric = FALSE`; `"ties_away"` is `"up"` with `symmetric = TRUE`. They exist because that translation is easy to get backwards, and because "up" means away from zero in Excel, Java, and Python --- but on the number line in scrutiny. The matching functions are `round_ties_up()`, `round_ties_down()`, `round_ties_away()`, and `round_ties_zero()`.

## Negative numbers: `symmetric`

`symmetric` matters whenever your data contain negative values: difference scores, z-scores, effect sizes, temperature scales.

By default (`symmetric = FALSE`), scrutiny reads "up" and "down" on the number line, so a negative tie goes up toward zero:

```{r}
round_up(-2.5)
round_down(-2.5)
```

With `symmetric = TRUE`, a negative number is rounded like its absolute value, so a tie moves *away* from zero:

```{r}
round_up(-2.5, symmetric = TRUE)
round_down(-2.5, symmetric = TRUE)
```

**Most statistical software does the second thing.** Excel, SPSS, SAS, and Matlab all round `-2.5` to `-3`; in IEEE 754 terms this is *roundTiesToAway*. Stata is the exception --- see the table below. So if you set `rounding` to a single direction on data with negative values, you almost always want `symmetric = TRUE` as well.

If you would rather not have to remember which way round that is, say it directly:

```{r}
round_ties_away(-2.5)   # Excel, SPSS, SAS, Matlab
round_ties_up(-2.5)     # Stata
```

This only ever affects ties, and only in negative numbers. And note that under the default `rounding = "up_or_down"`, the range spans both settings, so `symmetric` cannot change a verdict there.

## Rounding at 5 in other software

If the part cut off by rounding is exactly 5, how do these programs round --- and which scrutiny settings reconstruct them?

The `rounding` and `symmetric` columns are one way to say it; the `"ties_*"` equivalent is in brackets.

| Software | Function | `rounding` | `symmetric` | Note |
|:---|:---|:---|:---|:---|
| Excel | `ROUND()` | `"up"` (`"ties_away"`) | `TRUE` | [Absolute value is rounded, then the sign is reapplied](https://support.microsoft.com/en-us/office/round-a-number-f82b440b-689d-4221-b21e-38da99d33435) |
| Excel | `ROUNDUP()` | `"anti_trunc"` | --- | Away from zero, *not* ceiling |
| Excel | `ROUNDDOWN()` | `"trunc"` | --- | Toward zero |
| Google Sheets | `ROUND()`, `ROUNDUP()`, `ROUNDDOWN()` | as Excel | as Excel | Follows Excel |
| SPSS | [`RND()`](https://www.ibm.com/docs/en/spss-statistics/latest?topic=expressions-arithmetic-functions) | `"up"` (`"ties_away"`) | `TRUE` | Ties away from zero; the fuzz is user-settable (`SET FUZZBITS`) |
| SPSS | `TRUNC()` | `"trunc"` | --- | |
| SAS | [`ROUND()`](https://documentation.sas.com/doc/en/pgmsascdc/9.4_3.5/lefunctionsref/p0tj6cmga7p8qln1ejh6ebevm0c9.htm) | `"up"` (`"ties_away"`) | `TRUE` | "Rounds up the absolute value and restores the original sign", with fuzzing |
| SAS | `ROUNDE()`, `ROUNDZ()` | `"even"` | --- | `ROUNDE()` fuzzes, `ROUNDZ()` does not |
| Matlab | [`round()`](https://www.mathworks.com/help/matlab/ref/double.round.html) | `"up"` (`"ties_away"`) | `TRUE` | "Rounds away from zero to the nearest integer with larger magnitude"; a `TieBreaker` option overrides this |
| Stata | [`round()`](https://www.stata.com/manuals/fnmathematicalfunctions.pdf) | `"up"` (`"ties_up"`) | `FALSE` | "x is always rounded up to the larger value \[...\] `round(-4.5)` is −4" |
| R (>= 4.0.0) | [`round()`](https://rpubs.com/maechler/Rounding) | `"even"` | --- | Measures which neighbor is closer on the binary double |
| Python | [`round()`](https://docs.python.org/3/library/functions.html#round) | `"even"` | --- | Same idea as R |
| Python | [`decimal`](https://docs.python.org/3/library/decimal.html) `ROUND_HALF_UP` | `"up"` (`"ties_away"`) | `TRUE` | In this vocabulary, "up" means away from zero |
| NumPy | [`np.round()`](https://numpy.org/doc/stable/reference/generated/numpy.round.html) | `"even"` | --- | |

Three things this table is worth reading for:

**Stata is the odd one out.** Its manual is explicit that ties go to the larger value on the *number line*, so `round(4.5)` is 5 and `round(-4.5)` is −4. That is scrutiny's default, `symmetric = FALSE`. Every other program in the table mirrors negatives instead.

**"Rounding up" means two different things.** Excel's `ROUNDUP()` and Google Sheets' `ROUNDUP()` round *away from zero*: `ROUNDUP(-3.14159, 1)` is `-3.2`. That is not scrutiny's `"ceiling"`, which would give `-3.1`. For positive numbers the two coincide, which is why the confusion survives so easily. The same holds in reverse for `ROUNDDOWN()`, which is scrutiny's `"trunc"`, not `"floor"`.

scrutiny's `"anti_trunc"` is `ROUNDUP()`, and `"trunc"` is `ROUNDDOWN()`. The same pair appears as Java's `RoundingMode.UP` and `DOWN`, and as Python's `decimal.ROUND_UP` and `ROUND_DOWN`. Up to scrutiny 1.0.0 the match was inexact --- `round_anti_trunc()` moved a value away from zero even when it already sat on the rounding grid, so `round_anti_trunc(8.42, digits = 2)` was `8.43` where `ROUNDUP(8.42, 2)` is `8.42` --- but it no longer is.

**Several programs fuzz, and so does scrutiny.** SAS's `ROUND()` performs "extra computations, called fuzzing, to try to make the result agree with decimal arithmetic", and SPSS exposes the same idea as a user setting. scrutiny's rounding functions do this too, nudging a value by about `1.5e-9` so that `round_up(0.145, 2)` is `0.15` rather than `0.14`. This is deliberate: the software being reconstructed displayed `0.145` from the very same inexact double. See the `Floating-point tolerance` section of `round_up()`.

## Note on rounding to even

`rounding = "even"` calls base R's `round()`, and that is the *correct* reconstruction of R, Python, and NumPy: those systems break ties by the parity question as measured on the binary double, which is exactly what `round()` does.

What is unreliable is *predicting* the result from the decimal you see. Most decimal fractions have no exact binary representation, so a number that looks like a tie usually isn't one, and which way it goes depends on invisible mantissa digits. See this [explanation of `base::round()`](https://lhdjung.github.io/scrutiny/articles/rounding-in-depth.html#to-even-base-r), or a [more general article](https://docs.python.org/3/tutorial/floatingpoint.html#tut-fp-issues) about the limits of floating-point arithmetic.

For consistency testing this is handled rather than guessed: with `rounding = "even"`, `unround()` and the tests treat **both** bounds as inclusive, which can only widen a reconstructed range, never narrow it. Being too permissive is the safe direction for error detection.

## A note on scrutiny's names

scrutiny's `"up"` means *half*-up: numbers are rounded up only from 5, not from any lower value. In Excel and in ordinary speech, "rounding up" means what scrutiny calls `"ceiling"`. Read `"up"` and `"down"` as short for "up from 5" and "down from 5".

The `"ties_*"` options and the `round_ties_*()` functions were added because of this. They are not a rename: `round_up()` and the `"up"` string keep working exactly as before, and neither is deprecated. What the new names buy is that the procedure is stated rather than implied.

A note on why they are not called `round_half_up()` and friends, which would be the obvious choice given `janitor::round_half_up()`. That function is *roundTiesToAway* --- `janitor::round_half_up(-2.5)` is `-3` --- so a scrutiny function of the same name meaning ties-toward-`+Inf` would take the same arguments and give the opposite answer, with the attach order of `library()` deciding silently which one ran. Nor is janitor an outlier: in Java's `RoundingMode.HALF_UP`, Python's `decimal.ROUND_HALF_UP`, and .NET's `MidpointRounding.AwayFromZero`, "half up" already means away from zero, because "up" in that vocabulary means away from zero. `"ties_away"` says which one it is in any of these vocabularies.
