---
title: "Rounding options"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Rounding options}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message=FALSE}
library(scrutiny)
```

Many scrutiny functions have a `rounding` argument that controls how numbers are assumed to have been rounded. Specify `rounding` as any of the strings below.

## Which option to choose

For consistency testing, the default is almost always the right choice, and there is usually no need to identify the software behind a paper.

`rounding = "up_or_down"` counts a value set as consistent if the reported number matches either of the two results. Its range therefore spans every way of breaking a tie at 5: up, down, to even, away from zero, toward zero. It cannot reject a value set because you guessed the tie-breaking rule wrong. `rounding = "ceiling_or_floor"` is wider still and covers every rounding procedure whatsoever, including truncation and custom thresholds.

Committing to a single procedure (`"up"`, `"down"`, `"trunc"`, ...) only ever makes a test stricter --- `"even"` excepted, which comes to the same test as the default; see the note on rounding to even below. That is worth doing if you have positive evidence about how the numbers were produced. Without such evidence, it is a good way to get a false positive, so if in doubt, leave `rounding` alone.

## In scrutiny

These options return two rounded numbers for each input number. In consistency tests, this means counting a value set as consistent if the reported number matches either of the two rounded reconstructed numbers.

-   `"up_or_down"` (the default): one number rounded up from 5, one down from 5; in this order.
-   `"up_from_or_down_from"`: one number rounded up, one down --- but not from 5. Instead, specify `threshold` as the number from which the input should be rounded. Like `"up_or_down"`, it spans exactly one rounding step, wherever you put the threshold.
-   `"ceiling_or_floor"`: one number ceiled, one floored.

By contrast, these options only return one rounded number per input:

-   `"even"` rounds to the nearest number, breaking a tie toward the even one, using base R's own `round()`. See the note below.
-   `"up"` or `"down"` rounds in the specified direction, starting at 5.
-   `"up_from"` or `"down_from"` rounds in the specified direction, starting at some number other than 5. Specify this number via the `threshold` argument.
-   `"ceiling"` or `"floor"` always rounds to the next higher or lower decimal place, respectively.
-   `"trunc"` and `"anti_trunc"` always round towards zero or away from it.

Two arguments modify these procedures:

-   `symmetric` mirrors the whole procedure about zero, which for the tie-at-5 methods comes to deciding how ties in negative numbers are broken. The next section says why that matters.
-   `threshold` moves the point at which `"up_from"` and `"down_from"` switch direction. It means the same thing in both: `"up_from"` goes up when the part cut off by rounding is at least `threshold` tenths of a step, and `"down_from"` goes down when it is at most that many. See `round_up_from()`.

Finally, five options name a complete tie-breaking procedure, so that `symmetric` does not have to be given separately (and is ignored if it is):

-   `"ties_up"` sends a tie to the higher number, `"ties_down"` to the lower one.
-   `"ties_away"` sends a tie away from zero, `"ties_zero"` toward it.

A fifth name, `"ties_even"`, is another spelling of `"even"`. Rounding to even is its own mirror image, so `symmetric` was never relevant to it; the name is there because *roundTiesToEven* is IEEE 754's default direction, and a family whose point is naming the tie rule should not leave out the most common rule of all.

`"ties_up"` is `"up"` with `symmetric = FALSE`; `"ties_away"` is `"up"` with `symmetric = TRUE`. They exist because that translation is easy to get backwards, and because "up" means away from zero in Excel, Java, and Python, but on the number line in scrutiny. The matching functions are `round_ties_up()`, `round_ties_down()`, `round_ties_away()`, `round_ties_zero()`, and `round_ties_even()`.

Between them, these five names and `"ceiling"`, `"floor"`, and `"trunc"` cover every rounding direction IEEE 754 defines:

| IEEE 754 | scrutiny |
|:---|:---|
| *roundTiesToEven* (the standard's default) | `"even"`, `"ties_even"` |
| *roundTiesToAway* | `"ties_away"` (`"up"` with `symmetric = TRUE`) |
| *roundTowardPositive* | `"ceiling"` |
| *roundTowardNegative* | `"floor"` |
| *roundTowardZero* | `"trunc"` |

The standard has no name for `"ties_up"`, `"ties_zero"`, or `"anti_trunc"`, which are Stata's, Python's `ROUND_HALF_DOWN`, and Excel's `ROUNDUP()` respectively.

## Negative numbers: `symmetric`

`symmetric` matters whenever your data contain negative values: difference scores, z-scores, effect sizes, temperature scales.

By default (`symmetric = FALSE`), scrutiny reads "up" and "down" on the number line, so a negative tie goes up toward zero:

```{r}
round_up(-2.5)
round_down(-2.5)
```

With `symmetric = TRUE`, a negative number is rounded like its absolute value, so a tie moves *away* from zero:

```{r}
round_up(-2.5, symmetric = TRUE)
round_down(-2.5, symmetric = TRUE)
```

Most statistical software does the second thing. Excel, SPSS, SAS, and Matlab all round `-2.5` to `-3`; in IEEE 754 terms, this is *roundTiesToAway*. Stata is the exception, as the table below shows. So if you set `rounding` to a single direction on data with negative values, you almost always want `symmetric = TRUE` as well.

If you would rather not have to remember which way round that is, say it directly:

```{r}
round_ties_away(-2.5)   # Excel, SPSS, SAS, Matlab
round_ties_up(-2.5)     # Stata
```

This only ever affects negative numbers. With `"up"`, `"down"`, and `"up_or_down"` it only affects ties among those, because the value those procedures hesitate over is the midpoint, which sits at the same place either way. With `"up_from"` and `"down_from"` at some `threshold` other than 5, it affects more than ties, since the point they hesitate over is no longer the midpoint:

```{r}
round_up_from(-4.28, digits = 1, threshold = 9)
round_up_from(-4.28, digits = 1, threshold = 9, symmetric = TRUE)
```

Note also that under the default `rounding = "up_or_down"`, the range spans both settings, so `symmetric` cannot change a verdict there.

## Rounding at 5 in other software

If the part cut off by rounding is exactly 5, how do these programs round, and which scrutiny settings reconstruct them?

The `rounding` and `symmetric` columns are one way to say it; the `"ties_*"` equivalent is in brackets.

| Software | Function | `rounding` | `symmetric` | Note |
|:---|:---|:---|:---|:---|
| Excel | `ROUND()` | `"up"` (`"ties_away"`) | `TRUE` | [Absolute value is rounded, then the sign is reapplied](https://support.microsoft.com/en-US/Excel/round-a-number) |
| Excel | `ROUNDUP()` | `"anti_trunc"` | --- | Away from zero, *not* ceiling |
| Excel | `ROUNDDOWN()` | `"trunc"` | --- | Toward zero |
| Google Sheets | `ROUND()`, `ROUNDUP()`, `ROUNDDOWN()` | as Excel | as Excel | Follows Excel |
| SPSS | [`RND()`](https://www.ibm.com/docs/en/spss-statistics/latest?topic=expressions-arithmetic-functions) | `"up"` (`"ties_away"`) | `TRUE` | Ties away from zero; the fuzz is user-settable (`SET FUZZBITS`) |
| SPSS | `TRUNC()` | `"trunc"` | --- | |
| SAS | [`ROUND()`](https://documentation.sas.com/doc/en/pgmsascdc/9.4_3.5/lefunctionsref/p0tj6cmga7p8qln1ejh6ebevm0c9.htm) | `"up"` (`"ties_away"`) | `TRUE` | "Rounds up the absolute value and restores the original sign", with fuzzing |
| SAS | `ROUNDE()`, `ROUNDZ()` | `"even"` | --- | `ROUNDE()` fuzzes, `ROUNDZ()` does not |
| Matlab | [`round()`](https://www.mathworks.com/help/matlab/ref/double.round.html) | `"up"` (`"ties_away"`) | `TRUE` | "Rounds away from zero to the nearest integer with larger magnitude"; a `TieBreaker` option overrides this |
| Stata | [`round()`](https://www.stata.com/manuals/fnmathematicalfunctions.pdf) | `"up"` (`"ties_up"`) | `FALSE` | "x is always rounded up to the larger value \[...\] `round(-4.5)` is −4" |
| R (>= 4.0.0) | [`round()`](https://rpubs.com/maechler/Rounding) | `"even"` | --- | Measures which neighbor is closer on the binary double |
| Python | [`round()`](https://docs.python.org/3/library/functions.html#round) | `"even"` | --- | Same idea as R |
| Python | [`decimal`](https://docs.python.org/3/library/decimal.html) `ROUND_HALF_UP` | `"up"` (`"ties_away"`) | `TRUE` | In this vocabulary, "up" means away from zero |
| NumPy | [`np.round()`](https://numpy.org/doc/stable/reference/generated/numpy.round.html) | `"even"` | --- | |

A few points in the table are worth spelling out.

Stata is the odd one out. Its manual is explicit that ties go to the larger value on the *number line*, so `round(4.5)` is 5 and `round(-4.5)` is −4. That is scrutiny's default, `symmetric = FALSE`. Every other program in the table mirrors negatives instead.

"Rounding up" also means two different things. Excel's `ROUNDUP()` and Google Sheets' `ROUNDUP()` round *away from zero*: `ROUNDUP(-3.14159, 1)` is `-3.2`. That is not scrutiny's `"ceiling"`, which would give `-3.1`. For positive numbers the two coincide, which is how the confusion survives. The same holds in reverse for `ROUNDDOWN()`, which is scrutiny's `"trunc"`, not `"floor"`.

So scrutiny's `"anti_trunc"` is `ROUNDUP()`, and `"trunc"` is `ROUNDDOWN()`. The same pair appears as Java's `RoundingMode.UP` and `DOWN`, and as Python's `decimal.ROUND_UP` and `ROUND_DOWN`. Up to scrutiny 1.0.0 the match was inexact, because `round_anti_trunc()` moved a value away from zero even when it already sat on the rounding grid: `round_anti_trunc(8.42, digits = 2)` was `8.43` where `ROUNDUP(8.42, 2)` is `8.42`. It is exact now.

Several programs fuzz, and so does scrutiny. SAS's `ROUND()` performs "extra computations, called fuzzing, to try to make the result agree with decimal arithmetic", and SPSS exposes the same idea as a user setting. scrutiny's rounding functions do this too, nudging a value by about `1.5e-9` so that `round_up(0.145, 2)` is `0.15` rather than `0.14`. That is deliberate: the software being reconstructed displayed `0.145` from the very same inexact double. See the `Floating-point tolerance` section of `round_up()`.

## Note on rounding to even

`rounding = "even"` calls base R's `round()`, and that is the *correct* reconstruction of R, Python, and NumPy: those systems break ties by parity as measured on the binary double, which is what `round()` does.

What is unreliable is *predicting* the result from the decimal you see. Most decimal fractions have no exact binary representation, so a number that looks like a tie usually isn't one, and which way it goes depends on invisible mantissa digits. See this [explanation of `base::round()`](https://lhdjung.github.io/scrutiny/articles/rounding-in-depth.html#to-even-base-r), or a [more general article](https://docs.python.org/3/tutorial/floatingpoint.html#tut-fp-issues) about the limits of floating-point arithmetic.

For consistency testing, this is handled rather than guessed. With `rounding = "even"`, `unround()` and the tests treat both bounds as inclusive, which can only widen a reconstructed range, never narrow it. Being too permissive is the safe direction for error detection.

One consequence is worth knowing before you reach for it. Those bounds are the *same* bounds as `"up_or_down"`'s, so `rounding = "even"` gives GRIM, GRIMMER, and DEBIT exactly the verdicts the default gives --- it is the one setting that commits to a single procedure and tightens nothing. Knowing that a paper's numbers came out of R therefore buys no extra power here. It is not a gap to be closed: the tie that would do the tightening is not knowable from the decimal on the page.

## A note on scrutiny's names

scrutiny's `"up"` means *half*-up: numbers are rounded up only from 5, not from any lower value. In Excel and in ordinary speech, "rounding up" means what scrutiny calls `"ceiling"`. Read `"up"` and `"down"` as short for "up from 5" and "down from 5".

The `"ties_*"` options and the `round_ties_*()` functions were added because of this. They are not a rename: `round_up()` and the `"up"` string keep working exactly as before, and neither is deprecated. The new names simply state the procedure instead of implying it.

Why are they not called `round_half_up()` and friends, which would be the obvious choice given `janitor::round_half_up()`? That function is *roundTiesToAway* (`janitor::round_half_up(-2.5)` is `-3`), so a scrutiny function of the same name meaning ties toward `+Inf` would take the same arguments and give the opposite answer, with the attach order of `library()` deciding silently which one ran. And janitor is not unusual here: in Java's `RoundingMode.HALF_UP`, Python's `decimal.ROUND_HALF_UP`, and .NET's `MidpointRounding.AwayFromZero`, "half up" already means away from zero, because "up" in that vocabulary means away from zero. `"ties_away"` says which one it is in any of these vocabularies.
