---
title: "Data wrangling"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data wrangling}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
# Print enough significant digits that values like `10.46` are not
# abbreviated to `10.5` in the output tibbles:
options(pillar.sigfig = 4)
```

```{r setup, message=FALSE}
library(scrutiny)
```

In general, scrutiny's techniques for error detection are designed for a focus on the essential points, cutting out time-consuming repetition. There are some bottlenecks, however, such as entering decimal numbers as strings, or splitting strings that look like `"7.64 (1.5)"`.

This vignette shows how to save your time preparing data for error detection. It gives some general tips for these tasks, and then presents scrutiny's own specialized wrangling functions.

## Trailing zeros

### Motivation

One particular challenge when looking for numeric irregularities using R is that numeric values don't preserve trailing zeros. A mean reported as 5.30 is stored as `5.3`, and nothing in the data says which one it was. However, trailing zeros are as important to GRIM, GRIMMER, and DEBIT as any other trailing digits would be.

scrutiny's consistency tests solve this by asking you to state the number of decimal places explicitly: `grim_map()` has a `digits_x` argument, and `grimmer_map()` and `debit_map()` have `digits_x` and `digits_sd`. None of them has a default, so you can't forget. Values themselves stay numeric.

You may still want the trailing zeros back in the values --- for display, for export, or for a function outside of scrutiny that expects strings. That is what `restore_zeros()` is for, and it is covered below.

### Enter numbers as strings

#### Automated

Several R packages help to extract tables from PDF. I recommend tabulizer (not currently on CRAN; see [installation notes](https://stackoverflow.com/questions/70036429/having-issues-installing-tabulizer-package-in-r)). There are also the [pdftables](https://expersso.r-universe.dev/pdftables#) and [pdftools](https://ropensci.r-universe.dev/pdftools#) packages.

Using tabulizer requires Java to be installed. When it works well, tabulizer is a great tool for importing tables quickly and efficiently. It automatically captures values as strings, so trailing zeros are treated just like other digits.

However, tabulizer might sometimes struggle, especially with older PDF files. That is most likely the fault of the PDF format itself because it has no inbuilt support for tables, so any effort to extract them faces serious ambiguities. (See below, *Replace column names by row values*, for a solution to one such issue.)

If there are many tables in multiple files formatted in the same way, it can be useful to check if tabulizer reliably and accurately captures them. If it doesn't, you might have to use copy and paste.

#### With copy and paste

Perhaps not all R users know that RStudio features an option for multiple cursors. These are especially useful in conjunction with `tibble::tribble()`, which is available via scrutiny. Here's how to use multiple cursors in the present context:

1.  Copy a column of numbers from PDF, pressing and holding `Alt` on Windows or `option` on Mac. (This works at least in Adobe Acrobat.)
2.  Paste it into a `tribble()` call as below.
3.  Pressing and holding `Alt`/`option`, select all the copied numbers.
4.  Enter a comma at the end of each, for `tribble()`'s syntax.

You should then get something like this:

```{r}
flights1 <- tibble::tribble(
  ~x    ,
   8.97 ,
   2.61 ,
   7.26 ,
   3.64 ,
   9.26 ,
  10.46 ,
   7.39 ,
)
```

Take note of how many decimal places the numbers had in the PDF --- here, two. You will need that number for `digits_x` when you call `grim_map()`, and the data itself can no longer tell you.

All that's missing is the sample size. Add it either via another `tribble()` column as above or via `dplyr::mutate()`, which also comes with scrutiny:

```{r}
flights1 <- flights1 |>
  dplyr::mutate(n = 28)

flights1
```

### Restore trailing zeros

When you need the trailing zeros back in the values themselves --- not just stated via `digits_x` --- call `restore_zeros()`. It returns strings. Suppose all of the following numbers originally had one decimal place, but some no longer do:

```{r}
vec <- c(4, 6.9, 5, 4.2, 4.8, 7, 4)

vec |>
  decimal_places()
```

Now, get them back with `restore_zeros()`:

```{r}
vec |>
  restore_zeros()

vec |>
  restore_zeros() |>
  decimal_places()
```

This uses the default of going by the longest mantissa and padding the other strings with decimal zeros until they have that many decimal places. However, this is just a heuristic: The longest mantissa might itself have lost decimal places. Specify the `width` argument to explicitly state the desired mantissa length:

```{r}
vec |>
  restore_zeros(width = 2)

vec |>
  restore_zeros(width = 2) |>
  decimal_places()
```

A convenient way to restore trailing zeros in a data frame is `restore_zeros_df()`. By default, it operates on all columns that are coercible to numeric (factors don't count):

```{r}
iris <- tibble::as_tibble(iris)
iris |>
  restore_zeros_df(width = 3)
```

Specify columns mostly like you would in `dplyr::select()`:

```{r}
iris |>
  restore_zeros_df(starts_with("Sepal"), width = 3)
```

## Split strings by parentheses

### Basic usage

With summary data copied or extracted from PDF (see above), you might encounter values presented like `5.22 (0.73)`. Instead of manually teasing them apart, call `split_by_parens()`:

```{r}
flights2 <- tibble::tribble(
  ~drone        , ~selfpilot    ,
  "0.09 (0.21)" , "0.19 (0.13)" ,
  "0.19 (0.28)" , "0.53 (0.10)" ,
  "0.62 (0.16)" , "0.50 (0.11)" ,
  "0.15 (0.35)" , "0.57 (0.16)" ,
)

flights2 |>
  split_by_parens()
```

Optionally, transform these values into a more useful format:

```{r}
flights2 |>
  split_by_parens(transform = TRUE)
```

From here, you can call `debit_map()` almost right away (supposing you deal with binary distributions' means and standard deviations). The `x` and `sd` columns are strings because they were extracted from strings, so convert them to numeric first. They still have their trailing zeros at this point, which makes `decimal_places()` a reliable way to determine `digits_x` and `digits_sd`:

```{r}
flights2_split <- flights2 |>
  split_by_parens(transform = TRUE) |>
  dplyr::mutate(n = 80)

digits <- max(decimal_places(flights2_split$x))
digits

flights2_split |>
  dplyr::mutate(x = as.numeric(x), sd = as.numeric(sd)) |>
  debit_map(digits_x = digits, digits_sd = digits)
```

If your strings look like `"2.65 [0.27]"`, specify the `sep` argument as `"brackets"`. Likewise for `"2.65 {0.27}"` and `sep = "braces"`. What about other separators, as in `"2.65 <0.27>"`? Specify `sep` as those two substrings, like `sep = c("<", ">")`. In all of these cases, the output will be the same as the default would be if the strings were like `"2.65 (0.27)"`.

### Column name suffixes

The defaults for column name suffixes are (1) `"x"` for the part before the parentheses and (2) `"sd"` for the part inside of them. However, this won't fit for all data presented like `5.22 (0.73)`. Override the defaults by specifying `col1` and/or `col2`:

```{r}
flights2 |>
  split_by_parens(end1 = "beta", end2 = "se")
```

These suffixes become column names if `transform` is set to `TRUE`:

```{r}
flights2 |>
  split_by_parens(end1 = "beta", end2 = "se", transform = TRUE)
```

### Extract substrings from `before_parens()` and `inside_parens()`

There also are specific functions for extracting the parts of the individual string vectors before or inside the parentheses:

```{r}
flights3 <- flights2 |>
  dplyr::pull(selfpilot)

flights3

flights3 |>
  before_parens()

flights3 |>
  inside_parens()
```

## Replace column names by row values

When extracting tables from PDF with tabulizer, you might get data frames (converted from matrices) that have wrong, nondescript column names, while the correct column names are stored in one or more rows within the data frame itself.

I will first simulate the problem. `x` and `n` should be column names, but instead they are values in the first row:

```{r}
flights1_with_issues <- flights1 |>
  dplyr::mutate(x = as.character(x), n = as.character(n)) |>
  tibble::add_row(x = "x", n = "n", .before = 1)

colnames(flights1_with_issues) <- c("Var1", "Var2")

flights1_with_issues
```

To remedy the issue, call `row_to_colnames()` on the data frame. It will replace the column names by the values of one or more rows. The latter are specified by their position numbers as in `dplyr::slice()`. For these numbers, the default is `1` because the column names will often be stored in the first row, if at all. The specified row or rows are then dropped because they shouldn't have been rows in the first place.

With the above example:

```{r}
flights1_with_issues |>
  row_to_colnames()
```

Note that `n` is still a string vector, but this is true for all columns in tables extracted with tabulizer.
